---
title: "SPSS_FACTOR_ANALYSIS"
author: "Patience Heath"
date: "2025-08-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(psych)
a_questions = readRDS("A_questions.rds")

```


```{r}
a_questions = na.omit(a_questions)
factor_subset = a_questions[1:24]
r = cor(factor_subset)
determinant_value = det(r)
determinant_value
head(r)
```
### Iteration 1 determinant value 

```{r, echo = FALSE}
det_subset = factor_subset %>% select(-c("A3c" , "A3b", "A3m"))
r = cor(det_subset)
det(r)
```

### Iteration 2 determinant value 

```{r}
det_subset2 = det_subset %>% select(-c("A3j"))
r = cor(det_subset2)
det(r)
```

# At this point we now know what variables to use and will now move to SPSS to perform Factor Analysis. 
**SPSS also will run Bartletts and give us the KMO statistic** 


# Confirming Validity 

### Cronbachs alpha

All the alpha values are greater than 0.7 which confirms that the variables are internally consistent. 

Alpha values are displayed in order factor 1, factor 2, factor 3.

```{r}
factor_1 = a_questions %>% select(c("A3i", "A3l", "A3h", "A3k", "A3a", "A1b", "A4", "A3e", "A3d"))
factor_2 = a_questions %>% select(c("A2a", "A2b", "A2c", "A2e", "A2d", "A1a"))
factor_3 = a_questions %>% select(c("A3g", "A3f", "A3h"))


alpha(factor_1, check.keys = TRUE)
alpha(factor_2, check.keys = TRUE)
alpha(factor_3, check.keys = TRUE)
```


### Calculating AVE and CR scores in R based on SPSS output


### Function to get AVE and CR scores

Below is the function I made to get the average variance explained (AVE) and the composite reliability (CR) scores.

Some of the AVE are low (Factor 1 and Factor 3), but since the composite reliability is higher than 0.6 for all the convergent validity of the construct is adequate.

```{r, echo = FALSE}
ave_cr
```

### AVE and CR scores for factor 1

```{r, echo = FALSE}
lambda = c(0.827, 0.822, 0.751, 0.673, 0.641, 0.610, -0.561, -0.515, 0.505)
ave_cr(lambda)
```

### AVE and CR scores for factor 2

```{r, echo = FALSE}
lambda = c(-0.870, -0.846, -0.788, -0.768, -0.729, -0.637)
ave_cr(lambda)
```

### AVE and CR scores for factor 3

```{r, echo = FALSE}
lambda = c(0.794, 0.727, 0.611)
ave_cr(lambda)
```
## Summary Table of Reliability and Validity 

```{r}
Validity = data.frame(Metric = c("Cronbach's Alpha", "AVE", "Composite Reliability"),
                        Factor_1 = c(0.91, 0.44, 0.73),
                        Factor_2 = c(0.88, 0.60, 0.90),
                        Factor_3 = c(0.88, 0.51, 0.75))

Validity
```

